Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-05 doc/thesis/discussion.tex
Updated references

Our use of the \gls{BA} model makes several simplifying assumptions. First, we
assume homogeneity across the network with respect to node behaviour and
transmission risk. In reality, the attraction to high-degree nodes seems likely
to vary among individuals, as does their risk of transmitting or contracting
the virus. We have also assumed that all transmission risks are symmetric,
which is clearly false for all known modes of \gls{HIV} transmission, and that
infected individuals never recover but remain infectious indefinitely. These
assumptions were made for the purpose of keeping the model as simple as
possible, since this is the very first attempt to fit a contact network model
in a phylodynamic context. However, the Gillespie simulation algorithm built
into \software{netabc} can handle arbitrary transmission and removal rates
which need not be homogeneous across the network. Moreover, it is possible to
use kernel-\gls{ABC} to fit a model which relaxes some or all of these
assumptions, which may be a fruitful avenue for future investigation. Despite
the possible misspecification, our estimates of the power law exponent $\gamma$
were in line with those reported in the literature (\cref{tab:gamma}).

>>>>

Our use of the \gls{BA} model makes several simplifying assumptions. First, we
assume homogeneity across the network with respect to node behaviour and
transmission risk. In reality, the attraction to high-degree nodes seems likely
to vary among individuals, as does their risk of transmitting or contracting
the virus. We have also assumed that all transmission risks are symmetric,
which is clearly false for all known modes of \gls{HIV} transmission, and that
infected individuals never recover but remain infectious indefinitely. These
assumptions were made for the purpose of keeping the model as simple as
possible, since this is the very first attempt to fit a contact network model
in a phylodynamic context. However, the Gillespie simulation algorithm built
into \software{netabc} can handle arbitrary transmission and removal rates
which need not be homogeneous across the network. Moreover, it is possible to
use kernel-\gls{ABC} to fit a model which relaxes some or all of these
assumptions, which may be a fruitful avenue for future investigation. Despite
the possible misspecification, our estimates of the power law exponent $\gamma$
were within the range of values reported in the literature (\cref{tab:gamma}).

### 2016-05-05 doc/thesis/objective.tex
Updated references

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
schneeberger2004scale}, meaning that their degree distributions follow a power
law (although there has been some disagreement, see
\autocite{handcock2004likelihood, bansal2007individual}). Preferential
attachment has been postulated as a mechanism by which scale-free networks
could be generated~\autocite{barabasi1999emergence}. The \gls{BA}
model~\autocite{barabasi1999emergence} is one of the simplest preferential
attachment models, which makes it a natural choice to explore with our method.
The second aim of this work is to use simulations to investigate the parameters
of the \gls{BA} model, including whether they have a detectable impact on tree
shape, and whether they can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. At the time of this writing, there were $635400$ HIV
sequences publicly available in GenBank, annotated with 172 distinct countries
of origin. Since \gls{HIV} is almost always spread through either sexual
contact or sharing of injection drug supplies, the contact networks underlying
\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
be highly nonrandom. Moreover, since no cure yet exists, efforts to curtail the
progression of an epidemic have relied on preventing further transmissions
through measures such as \gls{tasp} and education leading to behaviour change.
The effectiveness of this type of intervention can vary significantly based on
the underlying structure of the network and the particular nodes to whom the
intervention is targeted~\autocite{little2014using,wang2015targeting}. Due to
this combination of data availability and potential public health impact,
\gls{HIV} is an obvious context in which our method could be applied.
Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
fit the \gls{BA} model to existing \gls{HIV} outbreaks.

>>>>

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
schneeberger2004scale, clemencon2015statistical}, meaning that their degree
distributions follow a power law (although there has been some disagreement,
see \autocite{handcock2004likelihood, bansal2007individual}). Preferential
attachment has been postulated as a mechanism by which scale-free networks
could be generated~\autocite{barabasi1999emergence}. The \gls{BA}
model~\autocite{barabasi1999emergence} is one of the simplest preferential
attachment models, which makes it a natural choice to explore with our method.
The second aim of this work is to use simulations to investigate the parameters
of the \gls{BA} model, including whether they have a detectable impact on tree
shape, and whether they can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. At the time of this writing, there were $635400$ HIV
sequences publicly available in GenBank, annotated with 172 distinct countries
of origin. Since \gls{HIV} is almost always spread through either sexual
contact or sharing of injection drug supplies, the contact networks underlying
\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
be highly nonrandom~\autocite{clemencon2015statistical}. Moreover, since no
cure yet exists, efforts to curtail the progression of an epidemic have relied
on preventing further transmissions through measures such as \gls{tasp} and
education leading to behaviour change. The effectiveness of this type of
intervention can vary significantly based on the underlying structure of the
network and the particular nodes to whom the intervention is
targeted~\autocite{little2014using,wang2015targeting}. Due to this combination
of data availability and potential public health impact, \gls{HIV} is an
obvious context in which our method could be applied. Therefore, the third and
final aim of this work is to apply kernel-\gls{ABC} to fit the \gls{BA} model
to existing \gls{HIV} outbreaks.

### 2016-05-05 doc/thesis/results.tex
Fix reference
Updated references

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1.1 
  [0.6 -
   1.16];
$I$,
  1109 
  [662 -
   4455];
$m$,
  3 
  [2 -
   5];
$N$,
  12673 
  [3733-
   14977].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.36 
  [0.01 -
   0.63];
$I$,
  2399 
  [1379 -
   3838];
$m$,
  3 
  [2 -
   5];
$N$,
  8830 
  [2836 -
   14720].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

@@

To make our analyses comparable to the existing network literature, we
estimated values of the power law exponent $\gamma$ for each of the datasets 
investigated. The results are summarized in \cref{tab:gamma}. All of the
estimated exponents were in the range $2 \leq \gamma \leq 4$ commonly reported
in the literature~\autocite{liljeros2001web, schneeberger2004scale,
colgate1989risk, brown2011transmission}.

>>>>

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1.1 
  [0.6 -
   1.16];
$I$,
  1137 
  [662 -
   4455];
$m$,
  3 
  [2 -
   5];
$N$,
  12785 
  [3861-
   14945].
The approximate posterior distributions for this simulation are shown in
\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
transmission tree in a peer-driven fashion, where the probability to sample a
node was twice as high if one of its peers had already been sampled. The
parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
$I$ = 2000. The estimated values were
$\alpha$, 
  0.37 
  [0.02 -
   0.63];
$I$,
  2352 
  [1423 -
   3811];
$m$,
  3 
  [2 -
   5];
$N$,
  8508 
  [2636 -
   14615].
The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
of these results were in line with estimates obtained on other simulated
datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
$\alpha$ was somewhat lower than typical.

@@

To make our analyses comparable to the existing network literature, we
estimated values of the power law exponent $\gamma$ for each of the datasets 
investigated. The results are summarized in \cref{tab:gamma}. All of the
estimated exponents were in the range $2 \leq \gamma \leq 2.5$.
